---
title: 'Regression Models Final Assignment : mtcars analysis'
author: "Alnour Ribault"
date: "22 février 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```
# Introduction
This document is my submission for the final assignment of the Regression Models course from the Coursera Data Science specialization by the John Hopkins University.

# Instructions

You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

* "Is an automatic or manual transmission better for MPG"
* "Quantify the MPG difference between automatic and manual transmissions"

Take the `mtcars` data set and write up an analysis to answer their question using regression models and exploratory data analyses.

# Author's note

Since we're doing many significance tests in this study, we have to correct for this. in order to avoid getting p-values by pure lucK. Since we estimate the number of significance tests in the study to be of the order of 10, the Bonferroni correction tells us to look for p-values under 0.005.

# Exploratory analysis

We first get a grip of the data by using basic R commands.

```{r basics, include=FALSE}
head(mtcars)
str(mtcars)
summary(mtcars)
```
It appears that some of the variables are naturally discrete. We thus convert them to factor variables.

```{r to_factor}
mtcars_fac <- mutate(mtcars, cyl = factor(cyl), 
                     vs = factor(vs, labels = c("V engine", "Straigth engine")),
                     am = factor(am, labels = c("Automatic", "Manual")), 
                     gear = factor(gear), 
                     carb = factor(carb))
```

We then plot a pair graph of the original `mtcars` data to get a grip of the correlation between the variables. `mpg` seems to decrease when `cyl`, `disp`, `hp`, `wt` increase, and seems to be higher among V engines than among Straigth engines, and higher among automatic transmission cars than among manual transmission cars as well.  
Some of those relations make sense : a heavier car will naturally use more gas, and a car designer will have to sacrifice some efficiency in order to achieve higher horsepower. The others, however, are more obscure.

Since we are especially interested in the relationship between variables `mpg` and `am`, we plot a boxplot of the value of `mpg` for automatic and manual transmission. It appears that cars with manual transmission have a notably higher mpg than those with automatic transmission. 

```{r firstFit}
fit <- lm(mpg ~ am, mtcars_fac)
```

Fitting a first model we find that the average value of `mpg` is `r round(coef(fit)[1],2)` for automatic cars and `r round(coef(fit)[2],2)` for manual cars. Both p-values are low enough for us to reject the null hypothesis that the actual coefficients are zero. However, the model's R-squared is only `r round(summary(fit)$r.squared,2)` which is not satisfying. We thus have to try adding other variables in order to explain `mpg`'s variance.

# Model Selection

As our first model is not satisfying, we will search for other significant variables in the modelling of `mpg`. We start with a model including all variables and use the AIC (Akaike information criterion) to eliminate variables down to a better model.

```{r model_selection, include = FALSE}
fit_all <- lm(mpg ~.,data=mtcars_fac)
fit_step <- step(fit_all)
```

This new model is not fully satisfying, however. The p-values for the `cyl8` coefficient is `r round(summary(fit_step)$coef[3,4],2)`, which is far from enough to attest for its significance. The p-value for the `am` coefficient is `r round(summary(fit_step)$coef[6,4],2)` which is not good enough either.

Since these coeffcients are not significative, let us fit two models, each with one of these variable dropped. We also fit a model where we drop both `cyl` and `am` variable and compare them to our previous model.

```{r noam}
fit_no_am <- lm(mpg ~ wt + hp + cyl, mtcars_fac)
fit_no_cyl <- lm(mpg ~ wt + hp + am, mtcars_fac)
fit_no_am_no_cyl <- lm(mpg ~ wt + hp, mtcars_fac)
anova(fit_step, fit_no_am)
anova(fit_step, fit_no_cyl)
anova(fit_step, fit_no_am_no_cyl)
```

The model without either `am` or `cyl` seems to be the only one improving on the previous one. It seems to be a pretty satisfying model since the p-values are better than previously, being under our Bonferroni threshold of 0.005!  
So the two remaining variables are horsepower and weight. But those two seem intuitively related : cars that have higher horsepower will be heavier. We thus try correcting for the interaction between the two variables.

```{r fina_fit}
fit_final <- lm(mpg ~ wt + hp + wt*hp, mtcars_fac)
anova(fit_no_am_no_cyl, fit_final)
```

The improvement obtained by adding the interaction term seems like a good idea. This model actually explains *almost* as much variance (R-squared = `r round(summary(fit_final)$r.squared,2)`) than the model containing all the variables (R-squared = `r round(summary(fit_all)$r.squared,2)`) which many less variables, *and* this time the coefficients of all the terms are highly significant.

# Residuals

We observe no particular pattern in the various plots of the residuals, and they seem to be normally distributed : our model presents no obvious weakness.

# Conclusion

Our study showed that, if the mtcars sample is representative, the transmission mode does not have a significant influence on the MPG, which is mainly explained by weight and horsepower. Therefore, the interaction between the transmission mode and the MPG is not quantifiable : the difference in MPG between the automatic cars and the manual cars is explained by their weight and horsepower. 

\pagebreak

# Appendices

## mtcars data

```{r basics_display, echo = FALSE}
head(mtcars)
str(mtcars)
summary(mtcars)
```

## Pair graph

```{r pairs}
pairs(mtcars, panel=panel.smooth, main="Pair graph for mtcars data")
```

## Boxplot

```{r boxplot}

g <- ggplot(mtcars_fac, aes(am, mpg)) +
    geom_boxplot(aes(fill = am)) +
    labs(title = "Value of mpg for automatic and manual transmissions", 
         x = "Transmission type", 
         y = "Miles/(US) gallon")
print(g)
```

# Models

## Simple Model

```{r simple_fit}
summary(fit)
```

## Model with all variables
```{r fit_all}
summary(fit_all)
```

## Model obtained with AIC
```{r fit_step}
summary(fit_step)
```

## Model with only `wt` and `hp`
```{r fit_no_am_no_cyl}
summary(fit_no_am_no_cyl)
```

## Final Model with `wt`, `hp` and their interaction
```{r fit_final}
summary(fit_final)
```
# Residuals

```{r residuals}
par(mfrow=c(2,2))
plot(fit_final)
```